%-----------------------------------------------------------------------
\subsection{Genetic Code Degeneracy (Biology)}
\label{sec:instance-genetic-code}
%-----------------------------------------------------------------------

The genetic code is \emph{degenerate}: 61 sense codons encode only 20 amino
acids, so multiple codons map to the same amino acid \citep{crick1966codon}.  For example, the
codons UCU and UCC both encode Serine.  This degeneracy is the biological
analogue of the Rashomon property: two distinct codons (configurations)
produce the same amino acid (observable) but carry incompatible nucleotide
information (explanations).

\paragraph{The explanation system.}
Define $\mathcal{S}_{\mathrm{genetic}} = (\Params, \sH, Y, \mathsf{obs},
\mathsf{exp}, \incomp)$ as follows:
\begin{itemize}
  \item $\Params = \sH = \{\text{Codon}\}$ (RNA codons);
  \item $Y = \{\text{AminoAcid}\}$ (the 20 amino acids);
  \item $\mathsf{obs} = \texttt{translate}$ (the codon-to-amino-acid map);
  \item $\mathsf{exp} = \mathrm{id}$ (the codon itself is the ``explanation''
    of the translation);
  \item $\incomp(c_1, c_2) \iff c_1 \ne c_2$.
\end{itemize}

\paragraph{The Rashomon property.}
The witnesses UCU and UCC satisfy
$\texttt{translate}(\text{UCU}) = \text{Ser} = \texttt{translate}(\text{UCC})$
and $\text{UCU} \ne \text{UCC}$.  The Rashomon property is constructively
witnessed with zero axioms.

\paragraph{The impossibility.}
By Theorem~\ref{thm:universal}, no method of inferring the codon from the
amino acid can simultaneously be faithful (report a codon consistent with the
actual one), stable (assign the same codon to all synonymous translations),
and decisive (distinguish every pair of distinct codons).  In biological terms:
knowing the protein sequence does not uniquely determine the DNA sequence.

\paragraph{Resolution: codon usage tables.}
The standard resolution in molecular biology is \emph{codon usage tables},
which report the frequency distribution over synonymous codons for each amino
acid.  This is the orbit average over the symmetry group of synonymous
substitutions: rather than committing to a single codon, one reports the
expected codon usage across the degenerate fiber.  The resolution sacrifices
decisiveness (no single codon is selected) to achieve stability and
faithfulness in expectation.

\paragraph{Lean formalization.}
The Lean~4 types are \texttt{Codon} (an inductive with constructors
\texttt{UCU} and \texttt{UCC}), \texttt{AminoAcid} (with constructor
\texttt{Ser}), and \texttt{geneticCodeSystem : ExplanationSystem Codon Codon
AminoAcid} with \texttt{observe := translate}, \texttt{explain := id}, and
\texttt{incompatible := ($\ne$)}.
The impossibility is \texttt{genetic\_code\_impossibility} in
\texttt{GeneticCode.lean}, derived with zero axioms.
