%-----------------------------------------------------------------------
\subsection{Microstate Degeneracy (Statistical Mechanics)}
\label{sec:instance-stat-mech}
%-----------------------------------------------------------------------

In statistical mechanics, a \emph{macrostate} (e.g., temperature, pressure)
is compatible with many \emph{microstates} (specific molecular configurations)
\citep{jaynes1957information}.
This coarse-graining---physically distinct configurations producing the same
observable---is the foundational example of underspecification in physics.
We formalize a minimal witness: a pair of binary coins.  This is not a
complete formalization of statistical mechanics but the simplest system
exhibiting the microstate/macrostate degeneracy that drives the impossibility
in real physical systems with $\sim 10^{23}$ particles.

\paragraph{The explanation system.}
Define $\mathcal{S}_{\mathrm{stat}} = (\Params, \sH, Y, \mathsf{obs},
\mathsf{exp}, \incomp)$ as follows:
\begin{itemize}
  \item $\Params = \sH = \texttt{CoinPair}$: a microstate is a pair of coins,
    each Heads or Tails;
  \item $Y = \mathbb{N}$: the macrostate, defined as the number of heads;
  \item $\mathsf{obs} = \texttt{numHeads}$;
  \item $\mathsf{exp} = \mathrm{id}$ (the full microstate is the
    ``explanation'');
  \item $\incomp(c_1, c_2) \iff c_1 \ne c_2$.
\end{itemize}

The permutation group $S_2$ acts on microstates by swapping the two coins:
$(H,T) \mapsto (T,H)$.  The macrostate is invariant under this action.

\paragraph{The Rashomon property.}
The witnesses are $(H,T)$ and $(T,H)$.  Both have macrostate $\texttt{numHeads} = 1$,
and they are distinct microstates.  Moreover, $(T,H) = \texttt{swapCoins}(H,T)$:
they are related by the $S_2$ permutation.  The Rashomon property is constructively
witnessed with zero axioms.

\paragraph{Universal permutation invariance.}
The theorem \texttt{swap\_preserves\_macrostate} is proved for \emph{all}
configurations, not just the two witnesses: for every $c \in \texttt{CoinPair}$,
$\texttt{numHeads}(\texttt{swapCoins}(c)) = \texttt{numHeads}(c)$.  The proof
proceeds by case analysis on all $2^2 = 4$ configurations.

\paragraph{The impossibility.}
By Theorem~\ref{thm:universal}, no description of a thermodynamic system can
simultaneously be faithful (specify a particular microstate), stable (assign the
same description to macrostate-equivalent configurations), and decisive
(distinguish every pair of distinct microstates).

\paragraph{Empirical illustration.}
For $N$ binary spins, the Rashomon entropy $S_R(k) = \ln \binom{N}{k}$ equals
the Boltzmann entropy.  Maximum faithfulness decays exponentially:
$1/\Omega(k) = e^{-S_R(k)}$.  At $k = N/2$ (maximum entropy),
$\Omega = \binom{50}{25} \approx 1.26 \times 10^{14}$, making faithful
microstate identification effectively impossible.

\paragraph{Resolution: the microcanonical ensemble.}
The microcanonical ensemble---the uniform distribution over microstates at fixed
macrostate---is structurally analogous to the $G$-invariant resolution.  For
permutation-symmetric systems, the connection is precise: the permutation group
acts on microstates, and the orbit average produces the macrostate.  Boltzmann's
program of replacing specific microstates with statistical distributions over
macrostates is an instance of the resolution strategy this paper formalizes.

\paragraph{Lean formalization.}
The Lean~4 types are \texttt{CoinPair} (a structure with fields
\texttt{coin1 coin2 : Bool}), \texttt{numHeads} (the observation map),
and \texttt{statMechSystem : ExplanationSystem CoinPair CoinPair Nat}.
The impossibility is \texttt{stat\_mech\_impossibility} in
\texttt{StatisticalMechanics.lean}, derived with zero axioms.
