%-----------------------------------------------------------------------
\subsection{Instance 8: LLM Self-Explanations Under Model Multiplicity}
\label{sec:instance-llm}
%-----------------------------------------------------------------------

Large language models are increasingly asked to explain their own
predictions: ``Classify this review as positive or negative and explain
which words drove your decision.''  We formalize such self-explanations
as citation-based summaries---template explanations citing the top-$k$
most-attended tokens---and show that they inherit the impossibility of
Theorem~\ref{thm:universal}.  Recent work on unfaithful chain-of-thought
reasoning \citep{turpin2024language} and measuring faithfulness of LLM
explanations \citep{lanham2023measuring} motivates the concern that LLM
self-explanations are unreliable even for models that agree on predictions.

\paragraph{The explanation system.}
We instantiate the abstract framework
(Definition~\ref{def:explanation-system}) as the 6-tuple
$\mathcal{S}_{\mathrm{llm}} = (\Params_{\mathrm{llm}},\,
\sH_{\mathrm{llm}},\,Y_{\mathrm{llm}},\,\mathsf{obs}_{\mathrm{llm}},\,
\mathsf{exp}_{\mathrm{llm}},\,\incomp_{\mathrm{llm}})$ where:
\begin{itemize}
  \item $\Params_{\mathrm{llm}}$ is the space of transformer weight
    configurations $\theta$ (attention heads, feed-forward layers, embeddings);
  \item $\sH_{\mathrm{llm}} = \binom{[T]}{k}$ is the set of size-$k$
    subsets of token positions, representing the $k$ tokens cited in the
    explanation (we use $k = 3$);
  \item $Y_{\mathrm{llm}}$ is the prediction space (e.g., binary sentiment
    labels);
  \item $\mathsf{obs}_{\mathrm{llm}}(\theta)$ is the predicted label for
    input sentence $x$;
  \item $\mathsf{exp}_{\mathrm{llm}}(\theta)$ extracts the top-$k$ content
    tokens by attention rollout importance as the ``cited evidence''; and
  \item $(S_1, S_2) \in \incomp_{\mathrm{llm}}$ when
    the highest-ranked token differs: $\max(S_1) \neq \max(S_2)$ under
    the attention ordering---i.e., the explanations disagree on the primary
    evidence token.
\end{itemize}

\paragraph{The Rashomon property.}
The attention rollout instability documented in
Section~\ref{sec:instance-attention} directly implies the Rashomon property
for $\mathcal{S}_{\mathrm{llm}}$: models with identical predictions but
different attention distributions cite different tokens as primary evidence.
\citet{ye2022unreliability} demonstrate empirically that LLM-generated
explanations are unreliable, changing substantially under minor prompt or
model variations while predictions remain stable.  \citet{turpin2024language}
show that chain-of-thought explanations can be systematically unfaithful
to the model's actual reasoning process.

\begin{proposition}[Instance 8: LLM Self-Explanation Impossibility]
\label{prop:llm-impossibility}
If $\mathcal{S}_{\mathrm{llm}}$ has the Rashomon property, then no
LLM self-explanation method can simultaneously be faithful, stable,
and decisive.
\end{proposition}

\begin{proof}
Direct application of Theorem~\ref{thm:universal} to
$\mathcal{S}_{\mathrm{llm}}$.
\end{proof}

\paragraph{Empirical validation.}
We validate Proposition~\ref{prop:llm-impossibility} using 10 perturbed
DistilBERT models ($\sigma = 0.01$, matching the attention experiment)
on 100 sentiment sentences.  Despite \textbf{100\% prediction agreement},
the \textbf{explanation flip rate} (fraction of model pairs citing a
different primary token) is \textbf{34.5\%} [95\% CI: 33.2--35.9\%],
and the mean pairwise \textbf{citation overlap} (Jaccard similarity of
top-3 token sets) is only \textbf{0.798} [CI: 0.791--0.806].
The \textbf{negative control} (top quartile of sentences by pairwise
Jaccard consensus, 26 sentences) achieves perfect citation overlap
(Jaccard = 1.000), confirming that instability is concentrated in
genuinely ambiguous sentences.
The \textbf{resolution test} (consensus top-3 from averaged attention
rollout) yields Jaccard overlap of \textbf{0.857} [CI: 0.842--0.871],
a \textbf{+5.8\%} improvement over pairwise overlap, confirming that
orbit averaging reduces citation instability.
Full results appear in Table~\ref{tab:llm_explanation_instability}.

\input{sections/table_llm_explanation}

\paragraph{Implications for regulatory requirements on LLM explanations.}
The EU AI Act \citep{euaiact2024} requires ``meaningful explanations'' for
high-risk AI systems, and practitioners increasingly rely on LLM
self-explanations to satisfy this requirement.
Proposition~\ref{prop:llm-impossibility} and the 34.5\% flip rate
demonstrate that such explanations are structurally unreliable: retraining
(or fine-tuning) the same architecture on the same data can change which
tokens the model cites as evidence, even when the prediction itself is
unchanged.  This is not a failure of any particular prompting strategy or
explanation method; it is a mathematical consequence of underspecification.
Regulators should require that LLM explanations be accompanied by
stability audits---e.g., reporting citation overlap across multiple model
instances---rather than treating a single model's self-explanation as
ground truth.
