% sections/related_work.tex — Related Work for the Universal Impossibility paper

\paragraph{Impossibility results in XAI.}
\citet{bilodeau2024impossibility} prove that no feature attribution satisfying
the Shapley axioms can be simultaneously faithful and stable when features are
collinear.  \citet{huang2024failings} show that SHAP values can be arbitrarily
misleading for Boolean classifiers.  \citet{rao2025limits} establish
algorithmic-information-theoretic limits on what any explanation method can
convey, and \citet{noguer2025mathematical} survey the mathematical foundations
of explainability more broadly.  Each of these results is proved in a
domain-specific language under domain-specific assumptions.  Our result
complements Bilodeau et al.\ (2024): they prove impossibility under Shapley
axioms for feature attributions; we prove impossibility under faithfulness,
stability, and decisiveness for arbitrary explanation systems.  The conditions
are incomparable---neither strictly implies the other---but both identify the
Rashomon property as the root cause.  Our contribution is a single abstract
theorem that unifies them as instances, showing the impossibility arises from
underspecification itself rather than from any particular axiom system.
\citet{krishna2022disagreement} document the complementary ``disagreement
problem'': different explanation methods produce different attributions for the
same model.  Our result addresses a distinct phenomenon---any single method
produces different explanations across equivalent models.

\paragraph{Underspecification and model multiplicity.}
\citet{damour2022underspecification} demonstrate that modern ML pipelines are
routinely underspecified: models with identical in-distribution performance
diverge on deployment-relevant properties.
\citet{fisher2019models} introduce \emph{model reliance} and
\emph{variable-importance clouds} that visualize how feature importance varies
across the set of near-optimal models.
\citet{rudin2024amazing} champion the \emph{Rashomon set}---the collection of
models with near-best loss---as a first-class object of study.
\citet{semenova2022existence} prove the existence of simpler models within
Rashomon sets, and \citet{marx2024uncertainty} develop uncertainty-aware
explainability that accounts for multiplicity.  Our framework formalizes the
bridge between underspecification and explanation impossibility: the Rashomon
property (Definition~\ref{def:rashomon}) is precisely the condition under which
equivalent hypotheses yield incompatible explanations.

\paragraph{Fairness impossibilities.}
\citet{chouldechova2017fair} proves that calibration and equal false-positive /
false-negative rates cannot simultaneously hold when base rates differ across
groups.  \citet{kleinberg2017inherent} establish an analogous incompatibility
for a broader class of fairness constraints.  We recast these as Instance~6 of
the universal impossibility: the explanation system maps each classifier to a
fairness certificate, and the Rashomon property holds whenever the underlying
population has unequal base rates.

\paragraph{Formal verification of impossibility theorems.}
\citet{nipkow2009social} formalized Arrow's impossibility theorem in
Isabelle/HOL, demonstrating that social-choice impossibilities can be
mechanically verified.  \citet{demoura2021lean4} introduced Lean~4 as a
combined theorem prover and programming language.
\citet{zhang2026statistical} recently formalized statistical learning theory in
Lean~4.  Our work provides, to our knowledge, the first Lean formalization of
an impossibility theorem in explainability, covering all nine instances and the
$G$-invariant resolution (75~files, 351~theorems, 72~axioms, 0~\texttt{sorry}).

\paragraph{Aggregation and resolution strategies.}
\citet{laberge2023partial} introduce a partial order on feature attributions
within the Rashomon set, extracting consensus rankings where they exist.
\citet{decker2024provably} optimize aggregation of feature attributions for
provably better explanations.
\citet{herren2023statistical} develop statistical inference methods for feature
rankings that are valid under model multiplicity.  We show that these diverse
strategies are all instances of the $G$-invariant resolution
(Section~\ref{sec:resolution}): they aggregate over the symmetry group of
equivalent hypotheses, restoring stability and faithfulness-in-expectation at
the cost of decisiveness.
