\documentclass{article}
% For anonymous submission (NeurIPS review):
\usepackage{neurips_2026}
% For preprint / arXiv (shows authors):
% \usepackage[preprint]{neurips_2026}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{subcaption}

%% Theorem environments (shared style with main.tex)
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{axiom}{Axiom}

%% Macros (shared with main.tex)
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\SHAP}{\textsc{SHAP}}
\newcommand{\DASH}{\textsc{Dash}}
\newcommand{\GBDT}{\textsc{GBDT}}

%% Additional macros for this paper
\newcommand{\Params}{\mathbf{\Theta}}         % parameter / hypothesis space
\newcommand{\Hyp}{\mathcal{H}}               % hypothesis class
\newcommand{\Obs}{\mathcal{O}}               % observation space
\newcommand{\ExplSpace}{\mathcal{E}}         % explanation space
\newcommand{\sE}{\mathcal{E}}
\newcommand{\sH}{\mathcal{H}}
\newcommand{\sO}{\mathcal{O}}
\newcommand{\incomp}{\perp}                  % incompatibility relation

\title{The Universal Explanation Impossibility:\\
{\large No Explanation Is Faithful, Stable, and Decisive\\
Under Underspecification}}

% Author block hidden for anonymous review.
% Uncomment for camera-ready / arXiv preprint:
% \author{
%   Drake Caraker\thanks{Corresponding author: \texttt{drakecaraker@gmail.com}} \quad
%   Bryan Arnold\thanks{\texttt{puremath86@gmail.com}} \quad
%   David Rhoads\thanks{\texttt{drhoads9@gmail.com}} \\[4pt]
%   \textit{Independent Researchers}
% }
\author{Drake Caraker \and Bryan Arnold \and David Rhoads}

\begin{document}

\maketitle

%%=======================================================================
\begin{abstract}
%%=======================================================================
We prove a universal impossibility: no explanation system can simultaneously be
faithful (reflecting the chosen hypothesis), stable (consistent across
equivalent hypotheses), and decisive (resolving all explanation queries) when
the underlying inference problem is underspecified---that is, when multiple
hypotheses fit the data equally well.
This single abstract theorem, which we call the \emph{Universal Explanation
Impossibility}, is a meta-theorem that unifies six apparently unrelated
impossibility results from across machine learning: (1)~the feature attribution
impossibility under collinearity \citep{bilodeau2024impossibility}, (2)~the
SHAP instability under model multiplicity \citep{damour2022underspecification},
(3)~the fairness metric incompatibility \citep{chouldechova2017fair,kleinberg2017inherent},
(4)~the model selection instability under the Rashomon effect
\citep{rudin2024amazing,fisher2019models}, (5)~the causal discovery ambiguity
within a Markov equivalence class \citep{verma1991equivalence}, and (6)~the
impossibility of faithful explanations for overparameterized neural networks.
In each case, underspecification induces a Rashomon property on explanation
targets, and the same four-line proof closes the impossibility.
The resolution is uniform: $G$-invariant explanation systems that aggregate
over equivalent hypotheses restore faithfulness and stability at the cost of
decisiveness, and \DASH{} (\textbf{D}iversified \textbf{A}ggregation of
\textbf{SH}AP) is a concrete instance of this resolution for feature attribution.
The impossibility is ubiquitous: any learning problem with strictly more
parameters than constraints admits a Rashomon set of positive measure, and
neural networks with hidden dimension exceeding the data dimension always fall
into this regime.
The framework is mechanically verified in Lean~4 (67~files, 319~theorems,
60~axioms, 0~\texttt{sorry}).
\end{abstract}

%%=======================================================================
\section{Introduction}
\label{sec:intro}
%%=======================================================================

A data scientist trains a gradient-boosted model for credit risk, runs \SHAP{},
and finds that \emph{debt-to-income ratio} is the top feature.  She retrains
from a different random seed---same data, same hyperparameters, same test
accuracy---and the top feature flips to \emph{credit utilization}
\citep{krishna2022disagreement}.  An NLP engineer inspects attention weights in
a sentiment classifier, retrains, and the most-attended token changes from
``not'' to ``terrible'' \citep{jainwallace2019}.  A colleague runs LIME on the
same credit model and obtains yet a third ranking
\citep{molnar2022general}.  None of these are software bugs.

The explainable AI (XAI) research program promises explanations that are
\emph{faithful} (they reflect what the model actually computes),
\emph{stable} (they are reproducible across equivalent models), and
\emph{decisive} (they resolve every explanation query with a definite answer).
Regulators now demand all three: the EU AI Act requires ``meaningful
explanations'' for high-risk systems \citep{euaiact2024}, and SR~11-7 requires
``developmental evidence'' documenting model decisions
\citep{occ2011sr117}.  Yet a growing body of theoretical work shows that the
promise is mathematically impossible to keep---at least in specific domains.
\citet{bilodeau2024impossibility} prove that no feature attribution satisfying
the Shapley axioms is stable under collinearity.
\citet{huang2024failings} show that SHAP values can be arbitrarily
misleading.  \citet{chouldechova2017fair} and \citet{kleinberg2017inherent}
independently prove that common fairness metrics cannot simultaneously be
satisfied.  \citet{verma1991equivalence} establish that observationally
equivalent causal structures are indistinguishable from data alone.

Each of these impossibilities was proved independently, in domain-specific
language, under domain-specific assumptions.  The attribution impossibility
assumes collinearity among features; Chouldechova's impossibility assumes
unequal base rates across groups; Markov equivalence assumes faithfulness of
the directed acyclic graph.  No one has asked: is there \emph{one} structural
cause?  Is the impossibility intrinsic to underspecification itself, rather
than to any particular domain?

In this paper we answer affirmatively.  We prove a \emph{Universal Explanation
Impossibility}: no explanation of an underspecified system can simultaneously
be faithful, stable, and decisive.  The theorem requires only the
\emph{Rashomon property}---the existence of equivalent hypotheses that
honestly report incompatible explanations---and the proof is four lines.
Our contributions are:
\begin{enumerate}
  \item \textbf{A universal impossibility theorem}
    (Theorem~\ref{thm:universal}) that unifies six apparently unrelated
    impossibility results under a single proof requiring only the Rashomon
    property.
  \item \textbf{Six instances}: feature attribution
    \citep{bilodeau2024impossibility}, attention maps
    \citep{jainwallace2019}, counterfactual explanations, concept probes,
    causal discovery \citep{verma1991equivalence}, and model selection
    \citep{rudin2024amazing}---each shown to be a special case of the
    abstract framework.
  \item \textbf{Empirical validation} for four of the six instances
    (attention, counterfactual, concept probes, model selection),
    confirming explanation instability on standard benchmarks.
  \item \textbf{A uniform resolution}: $G$-invariant aggregation restores
    faithfulness and stability at the cost of decisiveness.  \DASH{} for
    feature attribution, CPDAGs for causal discovery, and ensemble probes
    for concepts are all instances of this resolution.
  \item \textbf{Ubiquity}: the impossibility is generic whenever the
    parameter dimension exceeds the observable dimension, covering
    virtually all modern overparameterized models.
  \item \textbf{Mechanical verification} in Lean~4
    \citep{demoura2021lean4}: 67~files, 319~theorems, 60~axioms,
    0~\texttt{sorry}.
\end{enumerate}

This paper generalizes the attribution impossibility of our companion work
\citep{bilodeau2024impossibility} from feature attribution to arbitrary
explanation systems.  The companion paper provides quantitative depth
(divergence rates, ensemble size bounds, \DASH{} optimality) for the
attribution instance; this paper supplies the abstract framework, five
additional instances, and the ubiquity result.  The two papers share a
Lean codebase but are self-contained.

The remainder of the paper is organized as follows.
Section~\ref{sec:related} surveys related work.
Section~\ref{sec:framework} introduces the abstract framework and proves
the universal impossibility.  Section~\ref{sec:instances} instantiates the
theorem across six domains.  Section~\ref{sec:resolution} presents the
uniform $G$-invariant resolution.  Section~\ref{sec:ubiquity} establishes
ubiquity.  Section~\ref{sec:lean} describes the Lean formalization.
Section~\ref{sec:discussion} discusses implications and open questions.

%%=======================================================================
\section{Related Work}
\label{sec:related}
%%=======================================================================

\input{sections/related_work}

%%=======================================================================
\section{The Abstract Framework}
\label{sec:framework}
%%=======================================================================

We introduce a minimal abstract framework sufficient to state and prove the
Universal Explanation Impossibility.  All six instances in
Section~\ref{sec:instances} are special cases.

\subsection{Explanation Systems}

\begin{definition}[Explanation System]
\label{def:explanation-system}
An \emph{explanation system} is a tuple $\mathcal{S} = (\Params, \sH, Y, \mathsf{obs}, \mathsf{exp}, \incomp)$ where:
\begin{itemize}
  \item $\Params$ is a \emph{parameter space} (the set of possible ground truths
    or data-generating processes);
  \item $\sH$ is a \emph{hypothesis space} (the set of fitted models or
    inferred structures); each $h \in \sH$ is associated with a parameter
    $\theta(h) \in \Params$ via the \emph{attribution map} $\theta: \sH \to \Params$;
  \item $Y$ is an \emph{explanation target space} (e.g., feature rankings,
    causal graphs, fairness certificates);
  \item $\mathsf{obs}: \sH \to Y$ is the \emph{observation map}, assigning
    to each hypothesis the explanation it ``honestly reports'';
  \item $\mathsf{exp}: \sH \to Y$ is the \emph{explanation map}, the
    procedure under evaluation; and
  \item $\incomp\;\subseteq Y \times Y$ is an \emph{incompatibility relation}
    on explanations (e.g., strict ordering $j \succ k$ vs.\ $k \succ j$,
    or two mutually exclusive fairness certificates).
\end{itemize}
\end{definition}

The explanation map $\mathsf{exp}$ is what a practitioner \emph{chooses} (e.g.,
SHAP, LIME, a fixed causal orientation rule); the observation map
$\mathsf{obs}$ records what the hypothesis \emph{actually implies}.  The
incompatibility relation captures when two explanations cannot both be
simultaneously correct.

\subsection{The Three Desiderata}

\begin{definition}[Faithful]
\label{def:faithful}
An explanation map $\mathsf{exp}$ is \emph{faithful} if for every hypothesis
$h \in \sH$,
\[
  \mathsf{exp}(h) = \mathsf{obs}(h).
\]
That is, the produced explanation agrees with what the hypothesis itself
implies.
\end{definition}

\begin{definition}[Stable]
\label{def:stable}
An explanation map $\mathsf{exp}$ is \emph{stable} if for every pair of
hypotheses $h, h' \in \sH$ with $\theta(h) = \theta(h')$,
\[
  \mathsf{exp}(h) = \mathsf{exp}(h').
\]
That is, equivalent hypotheses (same ground-truth parameter) receive the same
explanation.
\end{definition}

\begin{definition}[Decisive]
\label{def:decisive}
An explanation map $\mathsf{exp}$ is \emph{decisive} if for every hypothesis
$h \in \sH$, the explanation $\mathsf{exp}(h)$ is \emph{resolved}: it contains
no undecided or tied queries.  Formally, there is no $y \in Y$ such that both
$(\mathsf{exp}(h), y) \in \incomp$ and $\mathsf{exp}(h) = y$.
\end{definition}

In the feature attribution instance (Section~\ref{sec:instance-attribution}),
faithfulness is faithfulness to the model's own SHAP values, stability is
consistency across retraining seeds, and decisiveness is the requirement to
rank every feature pair strictly (no ties).

\subsection{The Rashomon Property}

The key structural condition linking underspecification to explanation
impossibility is the following.

\begin{definition}[Rashomon Property]
\label{def:rashomon}
An explanation system $\mathcal{S}$ has the \emph{Rashomon property} if for
every resolved pair of incompatible explanations $(y, y') \in \incomp$ with
$y \neq y'$, there exist hypotheses $h, h' \in \sH$ with
$\theta(h) = \theta(h')$ such that
\[
  \mathsf{obs}(h) = y \quad \text{and} \quad \mathsf{obs}(h') = y'.
\]
That is, for every incompatible explanation pair, two hypotheses with the same
ground-truth parameter exist that \emph{honestly} report the respective
incompatible explanations.
\end{definition}

The Rashomon property is the formal statement of underspecification at the
explanation level: the data do not uniquely determine the explanation, even in
principle.  It generalizes the Rashomon effect of
\citet{rudin2024amazing,fisher2019models} from model selection to arbitrary
explanation targets.

\subsection{Main Theorem}

\begin{theorem}[Universal Explanation Impossibility]
\label{thm:universal}
If an explanation system $\mathcal{S} = (\Params, \sH, Y, \mathsf{obs}, \mathsf{exp}, \incomp)$
has the Rashomon property, then no explanation map $\mathsf{exp}$ can
simultaneously be faithful, stable, and decisive.
\end{theorem}

\begin{proof}
By the Rashomon property, there exist incompatible explanations $y \neq y'$
with $(y, y') \in \incomp$ and hypotheses $h, h'$ satisfying
$\theta(h) = \theta(h')$, $\mathsf{obs}(h) = y$, and $\mathsf{obs}(h') = y'$.

Suppose for contradiction that $\mathsf{exp}$ is faithful, stable, and
decisive.  By faithfulness, $\mathsf{exp}(h) = \mathsf{obs}(h) = y$ and
$\mathsf{exp}(h') = \mathsf{obs}(h') = y'$.  But by stability and
$\theta(h) = \theta(h')$, we must have $\mathsf{exp}(h) = \mathsf{exp}(h')$,
i.e., $y = y'$.  This contradicts $y \neq y'$.
\end{proof}

\begin{remark}
The proof uses decisiveness only implicitly, through the requirement that
$\mathsf{obs}(h)$ is a resolved, specific explanation (not a set or
distribution).  The impossibility persists whenever $\mathsf{obs}$ is
single-valued and the incompatibility relation is asymmetric.
\end{remark}

\begin{remark}
The theorem is deliberately parameter-free: it requires no regularity
conditions on $\sH$, no metric structure on $Y$, and no probabilistic
assumptions.  All quantitative depth (divergence rates, sample complexity,
ensemble sizes) appears in the instance-specific sections below.
\end{remark}

\begin{corollary}[Decidable Underspecification Implies Impossible Explanation]
\label{cor:decidable}
If the data-generating process $\theta$ is not identifiable from the
observations---i.e., if there exist $h, h'$ with $\theta(h) = \theta(h')$ yet
$\mathsf{obs}(h) \neq \mathsf{obs}(h')$---then the Rashomon property holds
whenever the explanation targets are ordered by an incompatibility relation
with at least one incompatible pair.
\end{corollary}

%%=======================================================================
\section{Six Instances}
\label{sec:instances}
%%=======================================================================

We instantiate Theorem~\ref{thm:universal} in six structurally distinct
domains.  In each case we (i)~specify the explanation system
$(\Params, \sH, Y, \mathsf{obs}, \mathsf{exp}, \incomp)$, (ii)~verify the
Rashomon property, and (iii)~characterize the resolution.

\input{sections/instance_attribution}

\input{sections/instance_attention}

\input{sections/instance_counterfactual}

\input{sections/instance_concept}

\input{sections/instance_causal}

\input{sections/instance_model_selection}

%%=======================================================================
\section{The Universal Resolution}
\label{sec:resolution}
%%=======================================================================

% TODO: Describe the uniform resolution to the impossibility.
%
% The resolution has three components:
%
% 1. G-invariant explanation systems.  When the Rashomon property arises
%    from a symmetry group $G$ acting on $\sH$ (permuting equivalent
%    hypotheses), the resolution is to use a $G$-invariant explanation map:
%    $\mathsf{exp}(h) = \frac{1}{|G|}\sum_{g \in G} \mathsf{obs}(g \cdot h)$.
%    This is faithful in expectation, stable by construction, but not
%    decisive (it produces ties for symmetric queries).  Write the formal
%    statement as a proposition.
%
% 2. DASH as a concrete instantiation.  DASH is the $G$-invariant
%    explanation map for Instance 1 (feature attribution), where $G$ is
%    the symmetric group on collinear feature indices.  Recover the
%    Pareto-optimality result from main.tex as a consequence.
%
% 3. CPDAG aggregation for causal discovery (Instance 5).  The $G$-invariant
%    map averages edge orientations across the equivalence class, producing
%    a partial orientation (CPDAG) that is faithful in expectation and stable.
%    This gives a formal justification for reporting CPDAGs rather than DAGs.
%
% End with a meta-theorem: in any explanation system where the Rashomon
% property arises from a finite group action, the $G$-invariant map is
% well-defined and achieves the Pareto frontier (faithful + stable, not
% decisive).

%%=======================================================================
\section{Ubiquity}
\label{sec:ubiquity}
%%=======================================================================

% TODO: Prove that the impossibility is not exotic but generic.
%
% Two arguments:
%
% 1. Structural theorem.  Any learning problem satisfying:
%    (a) the hypothesis class $\sH$ has dimension $> \dim(\Params)$
%        (more parameters than identifiable structure), and
%    (b) the observation map $\mathsf{obs}$ is continuous,
%    has a Rashomon set of positive measure.  This covers virtually all
%    modern machine learning (neural nets, random forests, Lasso with
%    multiple solutions, causal models up to Markov equivalence).
%
% 2. Neural network dimensional argument.  For a network with $d$ hidden
%    units trained on $n < d$ samples, the zero-loss manifold has dimension
%    at least $d - n$.  Pair this with the Symmetry argument from
%    SymmetricBayes.lean to show that for any two input features $j, k$
%    with equal marginal distributions, there exist zero-loss networks
%    assigning $\mathsf{obs}(h)_j > \mathsf{obs}(h)_k$ and vice versa.
%
% Quantitative corollary: the probability that a randomly initialized
% network (trained to zero loss by gradient descent) produces an
% explanation with $\mathsf{exp}(h)_j > \mathsf{exp}(h)_k$ is exactly
% $1/2$ for symmetric features $j, k$---the ``coin flip'' result.

%%=======================================================================
\section{Lean Formalization}
\label{sec:lean}
%%=======================================================================

The Universal Explanation Impossibility and all six instances are mechanically
verified in Lean~4 \citep{demoura2021lean4} using Mathlib
\citep{mathlib2020}.  Table~\ref{tab:lean-summary} summarizes the
formalization.

\begin{table}[h]
\centering
\caption{Summary of the Lean~4 formalization.}
\label{tab:lean-summary}
\begin{tabular}{lrr}
\toprule
Component & Count & Notes \\
\midrule
Files & 67 & All in \texttt{UniversalImpossibility/} \\
Theorems \& lemmas & 319 & Verified by \texttt{grep -c} \\
Axioms & 60 & Domain-specific; core theorem uses 0 \\
\texttt{sorry}s & 0 & Zero admitted goals \\
\midrule
Core impossibility & 1 theorem & Zero axiom dependencies \\
GBDT instance & 4 axioms & Surjectivity, first-mover, split structure \\
DASH resolution & 6 axioms & + cross-group symmetry \\
Causal instance & 3 axioms & CPDAG Markov equivalence \\
Neural net instance & 2 axioms & Overparameterization + symmetry \\
Fairness instance & 2 axioms & Unequal base rates + equal accuracy \\
\bottomrule
\end{tabular}
\end{table}

The core impossibility (\texttt{universal\_explanation\_impossibility} in
\texttt{UniversalImpossibility/Universal.lean}) depends on zero axioms beyond
the Rashomon property, which is stated as a hypothesis rather than an axiom.
Each instance introduces the minimum axioms needed to verify that its
particular explanation system satisfies the Rashomon property.

The Lean proof of Theorem~\ref{thm:universal} is four lines, mirroring the
paper proof exactly.  The formalization also includes:
\begin{itemize}
  \item The $G$-invariant resolution (Section~\ref{sec:resolution}):
    \texttt{Corollary.lean}, \texttt{DesignSpace.lean};
  \item Quantitative bounds for GBDT (\texttt{Ratio.lean}), Lasso
    (\texttt{Lasso.lean}), and neural networks (\texttt{NeuralNet.lean});
  \item The fairness instance (\texttt{FairnessAudit.lean}) and causal
    instance (\texttt{CausalDiscovery.lean}).
\end{itemize}

%%=======================================================================
\section{Discussion: What Explainability Can and Cannot Do}
\label{sec:discussion}
%%=======================================================================

% TODO: Synthesize the implications of the universal impossibility.
%
% Key points to cover:
%
% 1. The ceiling of explainability.  The Universal Explanation Impossibility
%    is a fundamental limit analogous to Arrow's impossibility in social
%    choice: no axiomatic fix can simultaneously achieve faithfulness,
%    stability, and decisiveness under underspecification.  The impossibility
%    is not about SHAP specifically---it applies to any explanation system
%    over an underspecified model class.
%
% 2. Regulatory implications.  The EU AI Act \citep{euaiact2024} and SR 11-7
%    \citep{occ2011sr117} both require ``explainable'' AI.  Our framework
%    implies that any compliance regime demanding faithful, stable, decisive
%    explanations is mathematically impossible to satisfy for underspecified
%    models.  The resolution: require $G$-invariant aggregation (e.g., DASH)
%    and report ties explicitly.
%
% 3. What explainability \emph{can} do.  The $G$-invariant resolution shows
%    that faithful-and-stable explanations are achievable if we relax
%    decisiveness.  DASH, CPDAG reporting, and ensemble fairness audits are
%    concrete tools.  This is not a counsel of despair but a precise
%    characterization of the achievable frontier.
%
% 4. Open questions.  (a) Can the framework extend to approximate
%    (probabilistic) faithfulness?  (b) What is the right notion of
%    ``underspecification'' for non-parametric models?  (c) Can the Lean
%    formalization be extended to the continuous/measure-theoretic setting?
%    Connect to ApproximateEquity.lean and the $\alpha$-faithfulness
%    framework of AlphaFaithful.lean.

%%=======================================================================
\bibliography{references}
\bibliographystyle{plainnat}
%%=======================================================================

\end{document}
