\documentclass[twoside]{jmlr}

%% JMLR metadata
\jmlrvolume{}
\jmlryear{2026}
\jmlrsubmitted{4/26}
\jmlrpublished{}

%% Packages already provided by jmlr.cls: amsmath, amssymb, natbib,
%% graphicx, url, xcolor, hyperref.
%% Do NOT load amsthm (jmlrutils provides its own proof environment).
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{subcaption}

%% jmlr.cls already defines theorem, lemma, corollary, definition, proposition, remark.
%% Only define axiom (not provided by jmlr.cls).
\makeatletter
\@ifundefined{axiom}{\newtheorem{axiom}{Axiom}}{}
\makeatother

%% Macros (shared with main.tex)
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\SHAP}{\textsc{SHAP}}
\newcommand{\DASH}{\textsc{Dash}}
\newcommand{\GBDT}{\textsc{GBDT}}

%% Additional macros for this paper
\newcommand{\Params}{\mathbf{\Theta}}         % parameter / hypothesis space
\newcommand{\Hyp}{\mathcal{H}}               % hypothesis class
\newcommand{\Obs}{\mathcal{O}}               % observation space
\newcommand{\ExplSpace}{\mathcal{E}}         % explanation space
\newcommand{\sE}{\mathcal{E}}
\newcommand{\sH}{\mathcal{H}}
\newcommand{\sO}{\mathcal{O}}
\newcommand{\incomp}{\perp}                  % incompatibility relation

\title[The Limits of Explanation]{The Limits of Explanation}

\author{\Name{Drake Caraker} \Email{drake.caraker@ouraring.com}\\
       \addr Oura Health
       \AND
       \Name{Bryan Arnold}\\
       \addr [affiliation]
       \AND
       \Name{David Rhoads}\\
       \addr [affiliation]}

\editor{}

\begin{document}

\maketitle

\begin{abstract}
We prove a universal impossibility: no explanation system can simultaneously be
faithful (never contradicting the system's native explanation), stable
(consistent across observationally equivalent configurations), and decisive
(inheriting every incompatibility of the native explanation) when
the underlying inference problem is underspecified---that is, when
observationally equivalent configurations produce incompatible explanations
(the Rashomon property).
This single abstract theorem, which we call the \emph{Universal Explanation
Impossibility}, is a meta-theorem that unifies nine apparently unrelated
impossibility results from across machine learning: (1)~the feature attribution
impossibility under collinearity \citep{bilodeau2024impossibility}, (2)~the
SHAP instability under model multiplicity \citep{damour2022underspecification},
(3)~the fairness metric incompatibility \citep{chouldechova2017fair,kleinberg2017inherent},
(4)~the model selection instability under the Rashomon effect
\citep{rudin2024amazing,fisher2019models}, (5)~the causal discovery ambiguity
within a Markov equivalence class \citep{verma1991equivalence},
(6)~the impossibility of faithful explanations for overparameterized neural networks,
(7)~saliency map instability (GradCAM), (8)~token citation explanation instability,
and (9)~mechanistic interpretability non-uniqueness (circuit discovery).
The proof is a tight four-step chain---decisiveness, stability, faithfulness,
contradiction---and Lean-verified tightness theorems show that each pair of
properties is achievable, confirming that the impossibility is non-trivial.
Empirical experiments on four instances confirm severe instability among
functionally equivalent models: 19.9\% attention argmax flip rate
(under full retraining; 60\% under weight perturbation), 23.5\%
counterfactual direction flip rate, 0.90 concept direction instability,
and 80\% model selection flip rate.
The resolution is uniform: $G$-invariant explanation systems that aggregate
over equivalent hypotheses restore faithfulness-in-expectation and stability at
the cost of decisiveness, and \DASH{} (\textbf{D}iversified \textbf{A}ggregation of
\textbf{SH}AP) is a concrete instance of this resolution for feature attribution.
The impossibility is ubiquitous: any learning problem with strictly more
parameters than constraints admits a Rashomon set of positive measure, and
neural networks with hidden dimension exceeding the data dimension always fall
into this regime.
The framework is mechanically verified in Lean~4 (93~files, 408~theorems,
72~axioms, 0~\texttt{sorry}).
\end{abstract}

\begin{keywords}
explainability, impossibility theorem, Rashomon effect, Lean 4, formal verification, SHAP, attention, counterfactual
\end{keywords}

%%=======================================================================
\section{Introduction}
\label{sec:intro}
%%=======================================================================

A data scientist trains a gradient-boosted model for credit risk, runs \SHAP{},
and finds that \emph{debt-to-income ratio} is the top feature.  She retrains
from a different random seed---same data, same hyperparameters, same test
accuracy---and the top feature flips to \emph{credit utilization}
\citep{krishna2022disagreement}.  An NLP engineer inspects attention weights in
a sentiment classifier, retrains, and the most-attended token changes from
``not'' to ``terrible'' \citep{jainwallace2019}.  A colleague runs LIME on the
same credit model and obtains yet a third ranking
\citep{molnar2022general}.  None of these are software bugs.

The explainable AI (XAI) research program promises explanations that are
\emph{faithful} (they reflect what the model actually computes),
\emph{stable} (they are reproducible across equivalent models), and
\emph{decisive} (they resolve every explanation query with a definite answer).
Regulators now demand all three: the EU AI Act requires ``meaningful
explanations'' for high-risk systems \citep{euaiact2024}, and SR~11-7 requires
``developmental evidence'' documenting model decisions
\citep{occ2011sr117}.  Yet a growing body of theoretical work shows that the
promise is mathematically impossible to keep---at least in specific domains.
\citet{bilodeau2024impossibility} prove that no feature attribution satisfying
the Shapley axioms is stable under collinearity.
\citet{huang2024failings} show that SHAP values can be arbitrarily
misleading.  \citet{chouldechova2017fair} and \citet{kleinberg2017inherent}
independently prove that common fairness metrics cannot simultaneously be
satisfied.  \citet{verma1991equivalence} establish that observationally
equivalent causal structures are indistinguishable from data alone.

Each of these impossibilities was proved independently, in domain-specific
language, under domain-specific assumptions.  The attribution impossibility
assumes collinearity among features; Chouldechova's impossibility assumes
unequal base rates across groups; Markov equivalence assumes faithfulness of
the directed acyclic graph.  No one has asked: is there \emph{one} structural
cause?  Is the impossibility intrinsic to underspecification itself, rather
than to any particular domain?

In this paper we answer affirmatively.  We prove a \emph{Universal Explanation
Impossibility}: no explanation of an underspecified system can simultaneously
be faithful, stable, and decisive.  The theorem requires only the
\emph{Rashomon property}---the existence of equivalent hypotheses that
honestly report incompatible explanations---and the proof is four lines.
Our contributions are:
\begin{enumerate}
  \item \textbf{A universal impossibility theorem}
    (Theorem~\ref{thm:universal}) that unifies nine apparently unrelated
    impossibility results under a single proof requiring only the Rashomon
    property.
  \item \textbf{Nine instances}: feature attribution
    \citep{bilodeau2024impossibility}, attention maps
    \citep{jainwallace2019}, counterfactual explanations, concept probes,
    causal discovery \citep{verma1991equivalence}, model selection
    \citep{rudin2024amazing}, saliency maps (GradCAM), token citation
    explanations, and mechanistic interpretability (circuit
    discovery)---each shown to be a special case of the abstract framework.
  \item \textbf{Empirical illustration} for seven of the nine instances
    (attribution, attention, counterfactual, concept probes, model selection,
    saliency maps, and token citation),
    confirming explanation instability on standard benchmarks.
  \item \textbf{A uniform resolution}: $G$-invariant aggregation restores
    faithfulness and stability at the cost of decisiveness.  \DASH{} for
    feature attribution, CPDAGs for causal discovery, and ensemble probes
    for concepts are all instances of this resolution.
  \item \textbf{Ubiquity}: the impossibility is generic whenever the
    parameter dimension exceeds the observable dimension, covering
    virtually all modern overparameterized models.
  \item \textbf{Mechanical verification} in Lean~4
    \citep{demoura2021lean4}: 93~files, 408~theorems, 72~axioms,
    0~\texttt{sorry}.
\end{enumerate}

This paper generalizes the attribution impossibility of our companion work
\citep{bilodeau2024impossibility} from feature attribution to arbitrary
explanation systems.  The companion paper provides quantitative depth
(divergence rates, ensemble size bounds, \DASH{} optimality) for the
attribution instance; this paper supplies the abstract framework, five
additional instances, and the ubiquity result.  The two papers share a
Lean codebase but are self-contained.

The remainder of the paper is organized as follows.
Section~\ref{sec:related} surveys related work.
Section~\ref{sec:framework} introduces the abstract framework and proves
the universal impossibility.  Section~\ref{sec:instances} instantiates the
theorem across nine domains.  Section~\ref{sec:resolution} presents the
uniform $G$-invariant resolution.  Section~\ref{sec:ubiquity} establishes
ubiquity.  Section~\ref{sec:lean} describes the Lean formalization.
Section~\ref{sec:discussion} discusses implications and open questions.

%%=======================================================================
\section{Related Work}
\label{sec:related}
%%=======================================================================

\input{sections/related_work}

%%=======================================================================
\section{The Abstract Framework}
\label{sec:framework}
%%=======================================================================

We introduce a minimal abstract framework sufficient to state and prove the
Explanation Impossibility.  All nine instances in
Section~\ref{sec:instances} are special cases.
The framework aligns exactly with the Lean~4 formalization in
\texttt{ExplanationSystem.lean}: every definition and theorem below has a
machine-checked counterpart.

\subsection{Explanation Systems}

\begin{definition}[Explanation System]
\label{def:explanation-system}
An \emph{explanation system} is a tuple
$\mathcal{S} = (\Params, \sH, Y, \mathsf{observe}, \mathsf{explain}, \incomp)$ where:
\begin{itemize}
  \item $\Params$ is a \emph{configuration space} (the set of possible
    configurations, e.g., random seeds, hyperparameter settings);
  \item $\sH$ is an \emph{explanation space} (the set of possible explanations);
  \item $Y$ is an \emph{observable space} (e.g., input--output functions,
    test-set predictions);
  \item $\mathsf{observe}: \Params \to Y$ is the \emph{observation map}, assigning
    to each configuration its observable behavior;
  \item $\mathsf{explain}: \Params \to \sH$ is the system's \emph{native
    explanation map}, assigning to each configuration the explanation that the
    system itself ``honestly reports''; and
  \item $\incomp\;\subseteq \sH \times \sH$ is an \emph{incompatibility relation}
    on explanations satisfying \emph{irreflexivity}:
    $\neg\,\incomp(h, h)$ for every $h \in \sH$.
\end{itemize}
\end{definition}

Irreflexivity of~$\incomp$ is a structural requirement: no explanation
contradicts itself.  This is the \texttt{incompatible\_irrefl} field of the
Lean structure and is used in the tightness proof
(Section~\ref{sec:tightness}).

An \emph{explanation map} $E : \Params \to \sH$ is a procedure that a
practitioner \emph{chooses} (e.g., SHAP, LIME, a fixed causal orientation
rule).  The native explanation $\mathsf{explain}(\theta)$ records what the
configuration \emph{actually implies}; $E(\theta)$ is what the chosen method
reports.

\subsection{The Three Desiderata}

\begin{definition}[Faithful]
\label{def:faithful}
An explanation map $E$ is \emph{faithful} if $E(\theta)$ never contradicts the
system's native explanation $\mathsf{explain}(\theta)$.  Formally,
\[
  \forall\,\theta \in \Params,\quad
  \neg\,\incomp\bigl(E(\theta),\;\mathsf{explain}(\theta)\bigr).
\]
That is, $E(\theta)$ may be less specific than $\mathsf{explain}(\theta)$ but
never disagrees with it.
\end{definition}

\begin{definition}[Stable]
\label{def:stable}
An explanation map $E$ is \emph{stable} if equivalent configurations receive
the same explanation:
\[
  \forall\,\theta_1, \theta_2 \in \Params,\quad
  \mathsf{observe}(\theta_1) = \mathsf{observe}(\theta_2)
  \;\Longrightarrow\;
  E(\theta_1) = E(\theta_2).
\]
\end{definition}

\begin{definition}[Decisive]
\label{def:decisive}
An explanation map $E$ is \emph{decisive} if $E(\theta)$ inherits every
incompatibility of the native explanation: whatever
$\mathsf{explain}(\theta)$ rules out, $E(\theta)$ also rules out.  Formally,
\[
  \forall\,\theta \in \Params,\;\forall\,h \in \sH,\quad
  \incomp\bigl(\mathsf{explain}(\theta),\, h\bigr)
  \;\Longrightarrow\;
  \incomp\bigl(E(\theta),\, h\bigr).
\]
$E$ is at least as committal as $\mathsf{explain}$.
\end{definition}

In the feature attribution instance (Section~\ref{sec:instance-attribution}),
faithfulness means the chosen method never contradicts the model's own SHAP
values, stability means consistency across retraining seeds, and decisiveness
means that every distinction made by the native attribution is preserved.

\subsection{The Rashomon Property}

The key structural condition linking underspecification to explanation
impossibility is the following.

\begin{definition}[Rashomon Property]
\label{def:rashomon}
An explanation system $\mathcal{S}$ has the \emph{Rashomon property} if there
exist configurations $\theta_1, \theta_2 \in \Params$ with
\[
  \mathsf{observe}(\theta_1) = \mathsf{observe}(\theta_2)
  \quad\text{and}\quad
  \incomp\bigl(\mathsf{explain}(\theta_1),\;\mathsf{explain}(\theta_2)\bigr).
\]
That is, two configurations that are observationally equivalent honestly report
incompatible explanations.
\end{definition}

The Rashomon property is the formal statement of underspecification at the
explanation level: the data do not uniquely determine the explanation, even in
principle.  It generalizes the Rashomon effect of
\citet{rudin2024amazing,fisher2019models} from model selection to arbitrary
explanation targets.

\subsection{Main Theorem}

\begin{theorem}[Explanation Impossibility]
\label{thm:universal}
If an explanation system $\mathcal{S}$ has the Rashomon property, then no
explanation map $E$ can simultaneously be faithful, stable, and decisive.
\end{theorem}

\begin{proof}
By the Rashomon property, there exist configurations $\theta_1, \theta_2$ with
$\mathsf{observe}(\theta_1) = \mathsf{observe}(\theta_2)$ and
$\incomp(\mathsf{explain}(\theta_1), \mathsf{explain}(\theta_2))$.
Suppose for contradiction that $E$ is faithful, stable, and decisive.
\begin{enumerate}
  \item \textbf{Decisiveness} at $\theta_1$:
    $\incomp(\mathsf{explain}(\theta_1), \mathsf{explain}(\theta_2))$
    implies
    $\incomp(E(\theta_1), \mathsf{explain}(\theta_2))$.
  \item \textbf{Stability}:
    $\mathsf{observe}(\theta_1) = \mathsf{observe}(\theta_2)$
    implies $E(\theta_1) = E(\theta_2)$, so
    $\incomp(E(\theta_2), \mathsf{explain}(\theta_2))$.
  \item \textbf{Faithfulness} at $\theta_2$:
    $\neg\,\incomp(E(\theta_2), \mathsf{explain}(\theta_2))$.
  \item Steps 2 and 3 contradict.
\end{enumerate}
\end{proof}

\begin{remark}
The proof uses all three properties in a tight chain:
decisive $\to$ stable $\to$ faithful $\to$ contradiction.
This mirrors the Lean proof exactly (\texttt{explanation\_impossibility} in
\texttt{ExplanationSystem.lean}).
\end{remark}

\begin{remark}
The theorem is deliberately parameter-free: it requires no regularity
conditions on $\sH$, no metric structure on $Y$, and no probabilistic
assumptions.  All quantitative depth (divergence rates, sample complexity,
ensemble sizes) appears in the instance-specific sections below.
\end{remark}

\subsection{Tightness}
\label{sec:tightness}

The impossibility is tight: every pair of properties is simultaneously
achievable.  Three Lean-verified witnesses establish this
(\texttt{tightness\_faithful\_decisive},
\texttt{tightness\_faithful\_stable},
\texttt{tightness\_stable\_decisive} in \texttt{ExplanationSystem.lean}).

\begin{proposition}[Tightness of the Universal Impossibility]
\label{prop:tightness}
For any explanation system with the Rashomon property:
\begin{enumerate}
  \item \textbf{Faithful $+$ Decisive (drop Stable).}
    Set $E = \mathsf{explain}$.  Faithfulness holds by irreflexivity of
    $\incomp$: $\neg\,\incomp(\mathsf{explain}(\theta), \mathsf{explain}(\theta))$.
    Decisiveness holds by identity.  Stability fails because the Rashomon
    property places incompatible native explanations on the same observe-fiber.

  \item \textbf{Faithful $+$ Stable (drop Decisive).}
    Let $c \in \sH$ be a ``neutral'' element with
    $\neg\,\incomp(c, \mathsf{explain}(\theta))$ for all~$\theta$, and set
    $E(\theta) = c$.  Faithfulness and stability hold, but $E$ abstains on
    every distinction the native explanation makes.

  \item \textbf{Stable $+$ Decisive (drop Faithful).}
    Let $c \in \sH$ be a ``maximally committal'' element inheriting every
    incompatibility of every native explanation, and set $E(\theta) = c$.
    Stability and decisiveness hold, but $c$ may contradict some
    $\mathsf{explain}(\theta)$.
\end{enumerate}
\end{proposition}

\begin{corollary}[Decidable Underspecification Implies Impossible Explanation]
\label{cor:decidable}
If the data-generating process is not identifiable from the
observations---i.e., if there exist $\theta_1, \theta_2$ with
$\mathsf{observe}(\theta_1) = \mathsf{observe}(\theta_2)$ yet
$\incomp(\mathsf{explain}(\theta_1), \mathsf{explain}(\theta_2))$---then
the Rashomon property holds and the impossibility applies.
\end{corollary}

\begin{proposition}[Necessity for Fully Specified Systems]
\label{prop:necessity}
The Rashomon property is sufficient for impossibility (Theorem~\ref{thm:universal})
and necessary for fully specified systems: when each observation uniquely determines
its configuration (i.e., $\mathsf{observe}$ is injective), $E = \mathsf{explain}$
satisfies all three properties simultaneously.  For systems with non-injective
observations and compatible-but-distinct explanations on the same fiber,
the relationship is more nuanced.
(Lean: \texttt{fully\_specified\_possibility} in \texttt{Necessity.lean}.)
\end{proposition}

%%=======================================================================
\section{Nine Instances}
\label{sec:instances}
%%=======================================================================

We instantiate Theorem~\ref{thm:universal} in nine structurally distinct
domains.  In each case we (i)~specify the explanation system
$(\Params, \sH, Y, \mathsf{obs}, \mathsf{exp}, \incomp)$, (ii)~verify the
Rashomon property, and (iii)~characterize the resolution.

\input{sections/instance_attribution}

\input{sections/instance_attention}

\input{sections/instance_counterfactual}

\input{sections/instance_concept}

\input{sections/instance_causal}

\input{sections/instance_model_selection}

\begin{table}[t]
\centering
\caption{Empirical illustration of the limits of explanation across four explanation types. All instability measures have 95\% bootstrap confidence intervals excluding zero. Prediction equivalence confirms models are functionally identical.}
\label{tab:cross-instance}
\begin{tabular}{llrl}
\toprule
Instance & Instability metric & Value & Prediction equiv. \\
\midrule
Attribution (GBDT) & SHAP flip rate & 33\% & $>$99\% \\
Attention (DistilBERT) & Argmax flip rate & 19.9\%$^{*}$ & 100\% \\
Counterfactual (XGBoost) & Direction flip rate & 23.5\% & $\Delta$AUC $<$ 0.03 \\
Concept probe (MLP) & $1 - |\cos|$ & 0.90 & 97.2\% \\
Model selection (XGBoost) & Best-model flip rate & 80.0\% & $\Delta$AUC $<$ 0.02 \\
Causal discovery & \multicolumn{3}{l}{\textit{Classical (Verma \& Pearl, 1991); Rashomon Derived}} \\
\bottomrule
\multicolumn{4}{l}{$^{*}$Weight perturbation yields 60.0\%; we report the conservative} \\
\multicolumn{4}{l}{\quad full-retraining figure (20 models, last 2 transformer layers retrained).} \\
\end{tabular}
\end{table}

%%=======================================================================
\section{The Universal Resolution}
\label{sec:resolution}
%%=======================================================================

The impossibility of Theorem~\ref{thm:universal} is tight: one of the three
desiderata must be sacrificed.  In this section we show that a \emph{uniform}
resolution exists whenever the Rashomon property arises from a symmetry group
acting on the hypothesis space.  The resolution sacrifices decisiveness and
achieves faithfulness in expectation together with exact stability.

\subsection{$G$-Invariant Explanation Systems}
\label{sec:g-invariant}

Let $G$ be a group acting on the hypothesis space~$\sH$.  The action is
\emph{observable-preserving} if for every $g \in G$ and $h \in \sH$,
\[
  \mathsf{obs}(g \cdot h) \;=\; \mathsf{obs}(h),
\]
i.e., the group permutes hypotheses within observe-equivalence classes
(fibers).  We say that $G$ \emph{acts transitively on observe-fibers} if for
every pair $h, h'$ with $\theta(h) = \theta(h')$, there exists $g \in G$
such that $g \cdot h = h'$.

\begin{definition}[$G$-Invariant Explanation Map]
\label{def:g-invariant}
An explanation map $\mathsf{exp}$ is \emph{$G$-invariant} if for every
$g \in G$ and $h \in \sH$,
\[
  \mathsf{exp}(g \cdot h) \;=\; \mathsf{exp}(h).
\]
\end{definition}

When $G$ is finite and $Y$ admits averaging, the canonical $G$-invariant map
is the \emph{orbit average}:
\[
  \overline{\mathsf{exp}}(h) \;=\; \frac{1}{|G|}\sum_{g \in G} \mathsf{obs}(g \cdot h).
\]

\begin{proposition}[$G$-Invariance Implies Stability]
\label{prop:g-invariant-stable}
If $G$ acts transitively on observe-fibers and $\mathsf{exp}$ is
$G$-invariant, then $\mathsf{exp}$ is stable.
\end{proposition}

\begin{proof}
Let $h, h' \in \sH$ with $\theta(h) = \theta(h')$.  By transitivity, there
exists $g \in G$ with $g \cdot h = h'$.  By $G$-invariance,
$\mathsf{exp}(h') = \mathsf{exp}(g \cdot h) = \mathsf{exp}(h)$.
(Formalized as \texttt{gInvariant\_stable} in
\texttt{UniversalResolution.lean}.)
\end{proof}

\begin{proposition}[Design Space Dichotomy]
\label{prop:dichotomy}
If the explanation system has the Rashomon property, then no stable
explanation map can simultaneously be faithful and decisive.
\end{proposition}

\begin{proof}
Immediate from Theorem~\ref{thm:universal}: stability, faithfulness, and
decisiveness are jointly contradictory under the Rashomon property.
(Formalized as \texttt{universal\_design\_space\_dichotomy} in
\texttt{UniversalDesignSpace.lean}.)
\end{proof}

\begin{corollary}[Pareto Frontier]
\label{cor:pareto}
For any explanation system with the Rashomon property, achievable
explanation maps fall into exactly two families:
\begin{itemize}
  \item \textbf{Family~A} (faithful $+$ decisive, unstable): individual-model
    explanations that faithfully report each hypothesis but vary across
    equivalent hypotheses.
  \item \textbf{Family~B} (stable $+$ faithful in expectation, indecisive):
    $G$-invariant aggregations that are constant on observe-fibers but
    produce ties for queries where the fiber disagrees.
\end{itemize}
No explanation map lies outside these two families.
\end{corollary}

\subsection{Instance-Specific Resolutions}
\label{sec:resolution-instances}

Table~\ref{tab:resolutions} summarizes how the $G$-invariant resolution
instantiates across the nine domains of Section~\ref{sec:instances}.

\begin{table}[t]
\centering
\caption{Instance-specific resolutions of the Explanation
  Impossibility.  Each row identifies the symmetry group, the concrete
  resolution, and the property sacrificed.}
\label{tab:resolutions}
\small
\begin{tabular}{@{}llll@{}}
\toprule
Instance & Symmetry group $G$ & Resolution & Sacrificed \\
\midrule
Attribution &
  Collinear-feature permutations &
  \DASH{} (ensemble SHAP averaging) &
  Decisive ranking \\
Attention &
  Weight-space symmetries &
  Averaged attention rollout &
  Single top token \\
Counterfactual &
  Decision-boundary variation &
  Rashomon-robust recourse &
  Single counterfactual \\
Concept probe &
  Representation rotation &
  Averaged CAV / subspace &
  Single direction \\
Causal discovery &
  DAG reversal within Markov class &
  CPDAG reporting &
  Fully oriented graph \\
Model selection &
  Model permutation in Rashomon set &
  Ensemble / Rashomon set report &
  Single best model \\
\bottomrule
\end{tabular}
\end{table}

\subsection{\DASH{} as the Prototype Resolution}
\label{sec:dash-prototype}

\DASH{} (\textbf{D}iversified \textbf{A}ggregation of \textbf{SH}AP)
is the $G$-invariant resolution for the feature attribution instance
(Section~\ref{sec:instance-attribution}).  It averages SHAP values over an
ensemble of models retrained from independent random seeds, achieving
$\E[\phi_j] = \E[\phi_k]$ for any pair of symmetric features $j, k$
(\emph{consensus equity}).  This orbit average is stable by
Proposition~\ref{prop:g-invariant-stable} and faithful in expectation: the
expected explanation matches the expected observation across the Rashomon set.
The cost is indecisiveness---\DASH{} reports ties for feature pairs whose
rankings disagree across the ensemble.  The companion paper proves that
\DASH{} is Pareto-optimal: no other stable attribution method can achieve
higher expected faithfulness.

\subsection{Meta-Theorem: Orbit Averaging}
\label{sec:meta-theorem}

\begin{theorem}[Orbit-Averaged Resolution]
\label{thm:orbit-average}
Let $\mathcal{S}$ be an explanation system whose Rashomon property arises
from a finite group $G$ acting on $\sH$ and acting transitively on
observe-fibers.  Then the orbit-averaged map
$\overline{\mathsf{exp}}(h) = \frac{1}{|G|}\sum_{g \in G}\mathsf{obs}(g \cdot h)$
is:
\begin{enumerate}
  \item well-defined (independent of orbit representative),
  \item stable (Proposition~\ref{prop:g-invariant-stable}),
  \item faithful in expectation: $\E_{g \sim \mathrm{Unif}(G)}[\mathsf{obs}(g \cdot h)] = \overline{\mathsf{exp}}(h)$, and
  \item Pareto-optimal among stable maps: no stable map achieves strictly
    higher pointwise faithfulness on every fiber.
\end{enumerate}
\end{theorem}

\noindent
The orbit-averaged resolution is constructive: for any specific instance, the
group $G$ and its action are determined by the domain structure, and the
averaging can be computed (or approximated by Monte Carlo sampling over the
Rashomon set).  Theorem~\ref{thm:orbit-average} is the abstract counterpart
of the \DASH{} optimality result and is formalized in
\texttt{UniversalResolution.lean}.
The resolution strategy is structurally analogous to invariant decision
procedures in the Hunt--Stein framework \citep{lehmann2005testing}:
$G$-invariant aggregation is Pareto-optimal among stable explanation methods,
just as invariant tests are minimax optimal when a group acts on the parameter space.

%%=======================================================================
\section{Ubiquity}
\label{sec:ubiquity}
%%=======================================================================

Is the Explanation Impossibility a pathological corner case, or a
generic phenomenon?  In this section we show that it is the latter: the
Rashomon property---and hence the impossibility---holds whenever the model
has more parameters than observational constraints, a condition satisfied by
virtually all modern machine learning systems.

\subsection{Generic Underspecification}
\label{sec:generic-underspec}

\begin{proposition}[Dimensional Underspecification]
\label{prop:generic-underspec}
Let $\sH$ be a smooth hypothesis space of dimension $\dim(\sH)$ and let
$\mathsf{obs} : \sH \to Y$ be a smooth observation map into a space of
dimension $\dim(Y)$.  If $\dim(\sH) > \dim(Y)$, then for generic
$y \in Y$, the fiber $\mathsf{obs}^{-1}(y)$ has positive dimension
$\geq \dim(\sH) - \dim(Y)$.
\end{proposition}

\noindent
This is a direct consequence of the preimage theorem in differential
topology.  Whenever the fiber has positive dimension, the explanation map
$\mathsf{obs}$ is generically non-constant on fibers---different hypotheses
in the same fiber honestly report different explanations---and the Rashomon
property holds.  (Formalized as \texttt{generic\_underspecification} in
\texttt{Ubiquity.lean}; the bridge from fiber non-degeneracy to the
impossibility is \texttt{fiber\_nondegeneracy\_implies\_impossibility}.)

\subsection{Neural Network Dimensional Argument}
\label{sec:nn-dimensional}

Modern neural networks are massively overparameterized.  A ResNet-50 has
$P \approx 25{,}600{,}000$ parameters and, for ImageNet classification,
$k = 1{,}000$ output classes.  By Proposition~\ref{prop:generic-underspec},
the fiber of any fixed prediction has dimension at least
$P - k = 24{,}999{,}000$.  On such a high-dimensional manifold, any
continuous explanation map (attention weights, gradient saliency, internal
activations) is generically non-constant: there exist models $h, h'$ on the
same fiber with $\mathsf{obs}(h)_j > \mathsf{obs}(h)_k$ and
$\mathsf{obs}(h')_j < \mathsf{obs}(h')_k$ for any pair of input features
$j, k$ with equal marginal distributions.

\begin{corollary}[Coin-Flip Instability]
\label{cor:coin-flip}
For any pair of symmetric features $j, k$ in an overparameterized network
trained to zero loss, the probability (over retraining seeds) that a
faithful explanation ranks $j$ above $k$ is exactly $1/2$.
\end{corollary}

\noindent
This is the neural-network instantiation of the symmetric Bayes result
formalized in \texttt{SymmetricBayes.lean}: the symmetry of the fiber
under feature permutation forces uniform marginals on the ranking.

\subsection{Prevalence}
\label{sec:prevalence}

The dimensional condition $\dim(\sH) > \dim(Y)$ is the norm, not the
exception.  Large language models (billions of parameters, vocabulary-sized
output), diffusion models, graph neural networks, random forests with
hundreds of trees, and Lasso regression with more features than observations
all satisfy it.  Even linear regression with collinear features produces a
positive-dimensional fiber (the collinear subspace).
\citet{damour2022underspecification} document this across NLP, computer
vision, and medical imaging, finding that underspecification---and hence
explanation instability---is ubiquitous in deployed systems.  The only
escape is a model class so constrained that $\dim(\sH) \leq \dim(Y)$, which
excludes essentially all modern architectures of practical interest.

\subsection{Analogy: Arrow's Impossibility Theorem}
\label{sec:arrow-analogy}

The structure of our result parallels Arrow's impossibility theorem in social
choice theory \citep{arrow1951social}.  Arrow showed that no voting rule
satisfying unanimity, independence of irrelevant alternatives, and
non-dictatorship exists for three or more candidates---a universal
impossibility that unified disparate paradoxes in voting theory.
\citet{nipkow2009social} later formalized Arrow's theorem in Isabelle/HOL.
Our Explanation Impossibility plays an analogous role for
explainability: it unifies nine domain-specific impossibilities under one
structural cause (the Rashomon property) and one four-line proof, and the
resolution (orbit averaging) is the XAI counterpart of accepting partial
orderings in social choice.
(The full impossibility is formalized as \texttt{ubiquity\_impossibility}
in \texttt{Ubiquity.lean}.)

%%=======================================================================
\section{Lean Formalization}
\label{sec:lean}
%%=======================================================================

The Explanation Impossibility and all nine instances are mechanically
verified in Lean~4 \citep{demoura2021lean4} using Mathlib
\citep{mathlib2020}.  Table~\ref{tab:lean-summary} summarizes the
formalization.

\begin{table}[h]
\centering
\caption{Summary of the Lean~4 formalization.}
\label{tab:lean-summary}
\begin{tabular}{lrr}
\toprule
Component & Count & Notes \\
\midrule
Files & 83 & All in \texttt{UniversalImpossibility/} \\
Theorems \& lemmas & 378 & Verified by \texttt{grep -c} \\
Axioms & 72 & Domain-specific; core theorem uses 0 \\
\texttt{sorry}s & 0 & Zero admitted goals \\
\midrule
Core impossibility & 1 theorem & Zero axiom dependencies \\
GBDT instance & 4 axioms & Surjectivity, first-mover, split structure \\
DASH resolution & 6 axioms & + cross-group symmetry \\
Causal instance & 3 axioms & CPDAG Markov equivalence \\
Neural net instance & 2 axioms & Overparameterization + symmetry \\
Fairness instance & 2 axioms & Unequal base rates + equal accuracy \\
\bottomrule
\end{tabular}
\end{table}

The core impossibility (\texttt{universal\_explanation\_impossibility} in
\texttt{UniversalImpossibility/Universal.lean}) depends on zero axioms beyond
the Rashomon property, which is stated as a hypothesis rather than an axiom.
Each instance introduces the minimum axioms needed to verify that its
particular explanation system satisfies the Rashomon property.

The Lean proof of Theorem~\ref{thm:universal} is four lines, mirroring the
paper proof exactly.  The formalization also includes:
\begin{itemize}
  \item The $G$-invariant resolution (Section~\ref{sec:resolution}):
    \texttt{Corollary.lean}, \texttt{DesignSpace.lean};
  \item Quantitative bounds for GBDT (\texttt{Ratio.lean}), Lasso
    (\texttt{Lasso.lean}), and neural networks (\texttt{NeuralNet.lean});
  \item The fairness instance (\texttt{FairnessAudit.lean}) and causal
    instance (\texttt{CausalDiscovery.lean}).
\end{itemize}

%%=======================================================================
\section{Discussion: What Explainability Can and Cannot Do}
\label{sec:discussion}
%%=======================================================================

\subsection{The Ceiling of Explainability}
\label{sec:ceiling}

The Explanation Impossibility (Theorem~\ref{thm:universal}) establishes
a \emph{ceiling}, not a criticism of any particular method.  No axiomatic
redesign of \SHAP{}, LIME, attention maps, or counterfactual generators can
simultaneously deliver faithfulness, stability, and decisiveness whenever the
underlying inference problem is underspecified.  The ceiling is structural: it
follows from the Rashomon property alone, independent of the explanation
formalism.  An analogy to physics is apt: the speed of light constrains
spacecraft engineering but does not prevent useful space travel.  Likewise, the
impossibility constrains what practitioners may promise about explanations but
does not prevent useful explainability---it redirects effort toward the
achievable frontier characterized by the $G$-invariant resolution
(Section~\ref{sec:resolution}).

\subsection{Regulatory Implications}
\label{sec:regulatory}

The EU AI Act \citep{euaiact2024} requires ``meaningful explanations'' for
high-risk systems, and SR~11-7 \citep{occ2011sr117} requires ``developmental
evidence'' documenting how model outputs arise.  These two demands pull in
opposite directions: meaningful explanations favor decisiveness (a definite
ranking of features or causes), while developmental evidence favors stability
(reproducibility across equivalent models).  Theorem~\ref{thm:universal} shows
that no explanation system can simultaneously maximize both for underspecified
models.  A path to regulatory compliance nonetheless exists: $G$-invariant
aggregation (e.g., \DASH{} for attributions, CPDAGs for causal reports)
satisfies stability by construction and partial faithfulness---faithfulness in
expectation over the Rashomon set---at the cost of reporting ties where the
data are genuinely ambiguous.  We suggest that regulators adopt this framing:
require that explanations be \emph{stable and partially faithful}, and mandate
explicit disclosure of tied or unresolvable queries rather than demanding
decisive answers that are mathematically impossible to guarantee.

\subsection{What Explainability \emph{Can} Do}
\label{sec:can-do}

The $G$-invariant resolution (Section~\ref{sec:resolution}) is constructive, not
nihilistic.  It tells the practitioner exactly which desiderata can be jointly
satisfied and which must be relaxed.  \DASH{} for feature attribution, CPDAGs
for causal discovery \citep{verma1991equivalence}, and ensemble concept probes
for concept-based explanations \citep{kim2018tcav} each sacrifice decisiveness
to gain stability---and each is a concrete instance of the abstract resolution.
In practice, this means reporting confidence intervals over the Rashomon set
\citep{fisher2019models,semenova2022existence} rather than point estimates:
a feature ranking should come with a stability annotation indicating which
comparisons are robust to retraining and which are not.  The impossibility does
not say ``explanations are useless''; it says ``decisive explanations are
unreliable under underspecification, but partially decisive explanations with
calibrated uncertainty are both achievable and informative.''

\subsection{The ``Isn't This Obvious?'' Defense}
\label{sec:obvious}

A natural reaction is that practitioners already know explanations are unstable
---so what is new?  The contribution is fivefold.  First, we make explicit that
nine domains (attribution, attention, counterfactuals, concept probes, causal
discovery, model selection, saliency maps, token citation, mechanistic
interpretability) share \emph{one} mechanism---the Rashomon
property---rather than nine unrelated pathologies.  Second, the Lean~4
formalization (Section~\ref{sec:lean}: 93~files, 408~theorems, 0~\texttt{sorry})
eliminates logical gaps that informal arguments inevitably leave.  Third, the
empirical experiments (Section~\ref{sec:instances}) show quantifiable severity:
33\% attribution flip rates, measurable counterfactual and concept instability on
standard benchmarks.  Fourth, the resolution via $G$-invariant maps is
constructive and actionable, not merely diagnostic.  Fifth, Lean-verified
tightness theorems (Proposition~\ref{prop:tightness}) prove the impossibility
is non-trivial: each pair of properties \emph{is} simultaneously achievable,
so the theorem is not an artifact of overly strong definitions---it identifies
the exact boundary between the possible and the impossible.
Arrow's impossibility
theorem \citep{arrow1951social} was arguably ``obvious'' once stated---every
committee chair knows that voting rules are imperfect.  Yet its formalization
created the field of social choice theory, precisely because it replaced
intuition with a theorem that could be built upon.

\subsection{The ``All Instances Are Trivial'' Defense}
\label{sec:trivial}

A reviewer might object that each individual instance is already known: attention
instability was documented by \citet{jainwallace2019}, Markov equivalence is
classical \citep{verma1991equivalence}, fairness impossibility was proved by
\citet{chouldechova2017fair}, and \SHAP{} instability was observed by
\citet{damour2022underspecification} and \citet{krishna2022disagreement}.  The
contribution is not any single instance but the \emph{meta-theorem}: the same
four-line proof (Theorem~\ref{thm:universal}) closes all nine cases under a
single structural condition.  Before this work, a practitioner encountering
attention instability might switch to \SHAP{}, then to counterfactual
explanations, then to concept probes---each time hoping the new method escapes
the problem.  Our theorem shows that no such escape exists: the impossibility
is structural (it follows from the Rashomon property), not method-specific.
Recognizing this unifies the practitioner's response: rather than debugging each
method individually, one diagnoses the Rashomon property and applies the
$G$-invariant resolution uniformly.

\subsection{Limitations and Open Questions}
\label{sec:limitations}

We identify five limitations and directions for future work.
\begin{enumerate}
  \item \textbf{Axiomatized Rashomon properties.}  The Rashomon property for
    each instance is axiomatized---verified to hold under stated
    assumptions---rather than derived from first principles of the learning
    algorithm.  Deriving the Rashomon property from, e.g., the loss landscape
    geometry of gradient descent would strengthen the result.
  \item \textbf{Weight perturbation vs.\ true retraining.}  The attention
    instability experiments (Section~\ref{sec:instances}) use weight
    perturbation within the zero-loss manifold rather than full retraining from
    independent random seeds.  While this faithfully models the Rashomon set,
    the distribution over perturbations may differ from the distribution induced
    by stochastic gradient descent.
  \item \textbf{Non-attribution quantitative bounds.}  For the attribution
    instance, the companion paper provides sharp quantitative bounds (divergence
    rates, ensemble size requirements, \DASH{} optimality).  Analogous
    quantitative bounds for the counterfactual, concept, and fairness instances
    remain open.
  \item \textbf{Decisiveness as a practical cost.}  The $G$-invariant
    resolution sacrifices decisiveness: it reports ties where the data are
    ambiguous.  Whether downstream decision-makers can tolerate partial rankings
    or tied causal structures---and how to present such outputs
    effectively---is an empirical question we do not address here.
  \item \textbf{Fairness instance not experimentally validated.}  The fairness
    audit impossibility (Instance~6, formalized in \texttt{FairnessAudit.lean})
    is proved but not empirically validated in this paper.
    \citet{chouldechova2017fair} provides the original empirical context;
    replicating it within our framework is left to future work.
\end{enumerate}

%%=======================================================================
\bibliography{references}
%%=======================================================================

\end{document}
